---
title: "COGO Bikes"
author: "Trevin Flickinger"
date: "10/30/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(lubridate)
library(ggthemes)
theme_set(theme_fivethirtyeight())
```

# Load the bike data adata

```{r, message = FALSE, warning = FALSE}
cogo <- read_csv("cogo.csv")
```

Check out trips over 1 minute vs total number of rows. I should probably be filtering out trips under 1 minute (not really a trip, mostly just checking out the bike or error).

```{r}
cogo %>%
  nrow()

cogo %>%
  mutate(trip_time = difftime(cogo$end_time, cogo$start_time, units = "mins")) %>%
  filter(trip_time > 1)
```


# Exploratory Data Discovery

When do people ride?

```{r Exploratory Data Discovery}
# Rides by Day of Week
cogo %>% 
  mutate(start_wday = lubridate::wday(start_time, label = TRUE)) %>%
  ggplot(aes(start_wday, fill = usertype)) +
  geom_bar() +
  labs(title = "COGO Rides by Day",
       subtitle = "Subscribers steady throughout week",
        x = "",
        y = "",
        caption = "@trevin_flick")


# Rides by time of day
cogo %>%
  ggplot(aes(format(as.POSIXct(start_time_round),format = "%H:%M"), fill = usertype)) +
  geom_bar() +
  labs(title = "COGO Rides by Time of Day",
       subtitle = "Most rides occur in evening",
        x = "",
        y = "",
        caption = "@trevin_flick") +
    theme(axis.text.x = element_text(size=8, angle=90, face="bold"),
        strip.text.x = element_text(size=12, face="bold"))

# Rides by time of day (weekdays only)
cogo %>%
  filter(start_wday %in% c('Mon','Tue','Wed','Thu','Fri')) %>%
  ggplot(aes(format(as.POSIXct(start_time_round),format = "%H:%M"), fill = usertype)) +
  geom_bar() +
  labs(title = "COGO Rides by Time of Day",
       subtitle = "Most rides occur in evening",
        x = "",
        y = "",
        caption = "@trevin_flick") +
    theme(axis.text.x = element_text(size=8, angle=90, face="bold"),
        strip.text.x = element_text(size=12, face="bold"))

# Rides by time of day
cogo_weather %>%
  filter(year == 2020 & month > 6) %>%
  ggplot(aes(time, fill = usertype)) +
  geom_bar() +
  facet_grid(~ month) +
  labs(title = "COGO Rides 2020",
       subtitle = "Last Four Months",
        x = "",
        y = "",
        caption = "@trevin_flick") +
  theme(axis.text.x = element_text(size=8, angle=90, face="bold"),
        strip.text.x = element_text(size=12, face="bold"))
```

Who is likely to ride?

```{r}
cogo_weather %>%
  mutate(age = year - birthyear) %>%
  ggplot(aes(age)) +
  geom_histogram() +
  theme_fivethirtyeight() +
  labs(title = "COGO Riders Skew Younger",
       subtitle = "Based on total trips by subscribers",
        x = "",
        y = "",
        caption = "@trevin_flick")

cogo_weather %>%
  drop_na(gender) %>%
  filter(year > 2013 & year < 2020) %>%
  ggplot(aes(year, fill = gender)) +
  geom_bar() +
  theme_fivethirtyeight() +
  labs(title = "COGO Riders Skew Male",
       subtitle = "Based on total trips by subscribers",
        x = "",
        y = "",
        caption = "@trevin_flick")
```

Where are people riding?

Are riders mostly doing A -> A trips or A -> B trips?

```{r}
# Rides by start station
cogo_weather %>%
  count(from_station_name, sort = TRUE)

# Top trips
cogo_weather %>%
  mutate(trip = paste(from_station_name, "->", to_station_name)) %>%
  count(trip, sort = TRUE)

# Top A -> B trips
cogo_weather %>%
  mutate(trip = paste(from_station_name, "->", to_station_name)) %>%
  filter(from_station_name != to_station_name) %>%
  count(trip, sort = TRUE)

cogo_trips <- as.tibble(cogo_weather) %>%
  mutate(trip = paste(from_station_name, "->", to_station_name),
         trip_type = if_else(from_station_name == to_station_name, "AA", "AB")) %>%
  filter(!is.na(from_station_name) & !is.na(to_station_name)) %>%
  select(from_station_name, to_station_name, trip, trip_type)

cogo_trips %>%
  count(trip_type, sort = TRUE)

cogo_trips %>%
  filter(is.na(trip_type))

colSums(is.na(cogo_trips))

cogo_weather %>%
  filter(is.na(from_station_name)) %>%
  View()


cogo_trips %>% 
  count(trip, trip_type) %>%
  top_n(25) %>%
  mutate(trip = fct_reorder(trip, n)) %>%
  ggplot(aes(trip, n, fill = trip_type)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c("#F8766D", "#999999"))  +
  labs(title = "Most popular trips",
       subtitle = "Downtown is a popular area for COGO riders",
        x = "",
        y = "",
        caption = "@trevin_flick")
  

# NA are expected, electric bikes do not need to be docked at a COGO station so start and end location may be missing

cogo_weather <- cogo_weather %>%
  separate(to_station_location, c("to_station_lat", "to_station_lon"), 
           ",", remove = FALSE)

cogo_weather <- cogo_weather %>%
  separate(from_station_location, c("from_station_lat", "from_station_lon"), 
           ",", remove = FALSE)

cogo_weather <- cogo_weather %>%
  mutate_at(c("to_station_lat", "to_station_lon", "from_station_lat", "from_station_lon"),
            as.numeric)

qmplot(to_station_lon, to_station_lat, data = cogo_weather, maptype = "toner-background", geom = c("point","density2d"),
       color = I("red"))
```

How long do people ride?

```{r}
cogo_weather %>%
  mutate(trip_time = difftime(cogo_weather$end_time, cogo_weather$start_time, units = "hours")) %>%
  filter(trip_time > 0 & trip_time < 3) %>%
  # filter(rideable_type == "electric_bike") %>%
  ggplot(aes(trip_time, temp)) +
  geom_point(alpha = 0.15)

cogo_weather %>%
  mutate(trip_time = difftime(cogo_weather$end_time, cogo_weather$start_time, units = "hours")) %>%
  filter(trip_time > 0 & trip_time < 3) %>%
  # filter(rideable_type == "electric_bike") %>%
  ggplot(aes(trip_time)) +
  geom_histogram() +
  geom_vline(xintercept = 0.5, color = "red", size = 1.5) +
  labs(title = "Duration of trips in hours",
       subtitle = "Most trips are under half hour",
        x = "",
        y = "",
        caption = "@trevin_flick")
```



# Number of Rides per day by Year

```{r}
cogo_weather %>%
  mutate(start_date = as.Date(start_time)) %>%
  ggplot(aes(x = start_date)) +
  geom_bar(fill = "#0099ff", color = "#0099ff") +
  theme_fivethirtyeight() +
  labs(title = "COGO Rides by Year",
       subtitle = "Number of rides for Columbus' bike share program",
        x = "",
        y = "",
        caption = "@trevin_flick") +
  facet_wrap(~year, ncol = 4, scales = "free_x") +
  theme(axis.text.x = element_text(size=8, angle=90, face="bold"),
        strip.text.x = element_text(size=12, face="bold"))

```

COGO introduced electric bikes to Columbus in the summer of 2020. Let's take a look at how users have responded to the new bike option.

```{r}
cogo_weather %>%
  mutate(start_date = as.Date(start_time)) %>%
  filter(year == 2020 & month >= 7) %>%
  ggplot(aes(x = start_date, fill = rideable_type)) +
  geom_bar() +
  theme_fivethirtyeight() +
  labs(title = "COGO Rides by Rideable Type",
       subtitle = "Number of rides for Columbus' bike share program",
       x = "",
       y = "",
       caption = "@trevin_flick") +
  theme(axis.text.x = element_text(size=8, angle=90, face="bold"),
        strip.text.x = element_text(size=12, face="bold"))
```


How does temp and weather effect rides?

```{r}
cogo_weather %>%
  count(start_time_round, temp) %>%
  ggplot(aes(temp, n)) +
  geom_point(alpha = 0.3) +
  theme_light() +
  labs(title = "Temperature has a big impact on rides",
       subtitle = "",
       x = "Temp (F)",
       y = "Rides per Hour",
       caption = "@trevin_flick")
```


# Modeling!!!!

```{r data prep}
library(tis)

# bring in weather data and join to bike data
weather <- read_csv("weather.csv")
cogo_weather <- inner_join(cogo, weather, by = c("start_time_round"="dt_est"))

# count number of rides per hour
cogo_count <- cogo_weather %>% 
  count(start_time_round, year, month, time, temp, pressure, humidity, wind_speed, name = "trips")

# data cleaning / feature engineering
cogo_count$start_date <- as.Date(cogo_count$start_time_round)
cogo_count$holiday <- isHoliday(cogo_count$start_date) * 1
cogo_count$day_of_week <- tis::dayOfWeek(cogo_count$start_date)
cogo_count$quarter <- tis::quarter(cogo_count$start_date)
cogo_count$hour_of_day <- as.numeric(cogo_count$time) / (60 * 60)

# select only variables of interest
cogo_count <- cogo_count %>%
  select(temp, pressure, humidity, wind_speed,
         year, quarter, month, day_of_week, hour_of_day, holiday, trips)
```


```{r}
library(tidymodels)
library(poissonreg)

set.seed(1)

month_split <- cogo_count %>%
  initial_split(strata = month)

cogo_train <- training(month_split)
cogo_test <- testing(month_split)

# linear regression model
lm_fit <- linear_reg() %>%
  set_engine(engine = 'lm') %>%
  fit(trips ~ .,
    data = cogo_train
  )

# poisson regression model
p_fit <- poisson_reg() %>%
  set_engine(engine = 'glm') %>%
  fit(trips ~ .,
    data = cogo_train
  )

# random forest
rf_fit <- rand_forest(mode = "regression") %>%
  set_engine(engine = "ranger") %>%
  fit(trips ~ .,
    data = cogo_train
  )
```


```{r}
# evaluate models
# training data
results_train <- lm_fit %>%
  predict(new_data = cogo_train) %>%
  mutate(
    truth = cogo_train$trips,
    model = "lm"
  ) %>%
  bind_rows(p_fit %>%
    predict(new_data = cogo_train) %>%
    mutate(
      truth = cogo_train$trips,
      model = "poisson"
    )) %>%
  bind_rows(rf_fit %>%
    predict(new_data = cogo_train) %>%
    mutate(
      truth = cogo_train$trips,
      model = "rf"
    ))

results_train %>%
  group_by(model) %>%
  rmse(truth = truth, estimate = .pred)
```




```{r}
# evaluate testing data

results_test <- lm_fit %>%
  predict(new_data = cogo_test) %>%
  mutate(
    truth = cogo_test$trips,
    model = "lm"
  ) %>%
  bind_rows(p_fit %>%
    predict(new_data = cogo_test) %>%
    mutate(
      truth = cogo_test$trips,
      model = "poisson"
    )) %>%
  bind_rows(rf_fit %>%
    predict(new_data = cogo_test) %>%
    mutate(
      truth = cogo_test$trips,
      model = "rf"
    ))

results_test %>%
  group_by(model) %>%
  rmse(truth = truth, estimate = .pred)

```





  
  
  
  
  
  
  
  
  
  
  